---
title: "Modern Applied Data Analysis Project"
subtitle: ""
author: Aidan Troha & Vijay Panthayi
date: "`r Sys.Date()`"
format:
  docx:
    toc: false
    number-sections: true
    highlight-style: github
bibliography: ../dataanalysis_template_references.bib
csl: ../apa.csl
---

```{r setup, include=FALSE}
library(readr)
library(tidyverse)
```

# PART 1:

## Briefly describe what the data is?
This data set is a cumulative data report on the number of deaths, by state (including Puerto Rico), due to COVID-19, Pneumonia, Influenza, a combination of two of those causes, or all three. The data set divides the data by state, then by sex, then by age group. 

## How it was collected?
Individual state data was reported to NCHS during the collection period. 

## Where you will get (or got) it from?
This data set was obtained directly from the CDC’s online data catalog at data.cdc.gov.
Data: https://data.cdc.gov/NCHS/Provisional-COVID-19-Deaths-by-Sex-and-Age/9bhg-hcku

## How many observations do you have, what was measured?
There are 115,668 observations in this data set. The variables that were measured were the state where the death occurred, the sex of the person who died, and the age group that person falls into. In addition, the data set measured the number of deaths attributed to each of the following: COVID-19, Pneumonia, both COVID-19 and Pneumonia, Influenza, all three, and the total deaths. The reason there are so many observations is because the data set is split into three groups: by total, by year, and by year and month. The group name describes how many observations per state were represented (by total would be a cumulative summary of the by year/by year and month sections).

## Anything else important to report about the data?
According to the data set, many observations were made null if the total number of deaths during the collection period was betwene 1-9. The CDC notes this is due to being in accordance with NCHS confidentiality standards. Also, for the purpose of our project and report, we will only be using the data which summarizes the total values for the year and each month (only using observations 1-2755; the by total group). This is for the purpose of simplicity while also allowing us to work with a smaller (but still sufficient) data set size.

## Question we want to answer with this data?
We are using this data to answer what factors resulted in the most deaths to COVID, Influenza, and/or Pneumonia. In addition, we can use the fact that the data is separated by state to also analyze geographical influence/correlations as well.

## Outcome of interest?
A statistical representation/test on the significance of each variable on the number of deaths due to COVID, Influenza, and/or Pneumonia.

## Predictors?
The predictors we are focusing on are the following: geographical location, age, and sex.

## Relations/Patterns?
We are looking to see if there are any relationships in the data.

## Data Analysis Plan?
We plan to test the data using some sort of regression analysis or odds ratio to determine the correlation between predictors and outcome.


```{r}
# Loads the data into R. The original data is in the "CSV" format
raw_data <- read_csv("../../data/raw_data/Provisional_COVID-19_Deaths_by_Sex_and_Age.csv")
# Provides an overview of the data and data types in the data frame.
glimpse(raw_data)
summary(raw_data)
```

# Part 2
```{r}
# Includes only the data we intend to use for our analysis.
useable <- raw_data %>%
# Identifies total values for each 
                    filter(Group == "By Total") %>%
# rids the dataset of "Notes" section
                    select(!c(Footnote,`Data As Of`,`Start Date`,`End Date`,Group,Year,Month))
glimpse(useable)
summary(useable)
```


# References



