---
title: "Modern Applied Data Analysis Project"
subtitle: ""
author: Aidan Troha & Vijay Panthayi
date: "`r Sys.Date()`"
format:
  docx:
    toc: false
    number-sections: true
    highlight-style: github
bibliography: ../dataanalysis_template_references.bib
csl: ../apa.csl
---

```{r setup, include=FALSE}
# Load necessary packages
library(readr)
library(tidyverse)
```

# PART 1(*has been placed in "proposal.Rmd" file in products folder*):

## Briefly describe what the data is?
This data set is a cumulative data report on the number of deaths, by state (including Puerto Rico), due to COVID-19, Pneumonia, Influenza, a combination of two of those causes, or all three. The data set divides the data by state, then by sex, then by age group. 

## How it was collected?
Individual state data was reported to NCHS during the collection period. 

## Where you will get (or got) it from?
This data set was obtained directly from the CDC’s online data catalog at data.cdc.gov.
Data: https://data.cdc.gov/NCHS/Provisional-COVID-19-Deaths-by-Sex-and-Age/9bhg-hcku

## How many observations do you have, what was measured?
There are 115,668 observations in this data set. The variables that were measured were the state where the death occurred, the sex of the person who died, and the age group that person falls into. In addition, the data set measured the number of deaths attributed to each of the following: COVID-19, Pneumonia, both COVID-19 and Pneumonia, Influenza, all three, and the total deaths. The reason there are so many observations is because the data set is split into three groups: by total, by year, and by year and month. The group name describes how many observations per state were represented (by total would be a cumulative summary of the by year/by year and month sections).

## Anything else important to report about the data?
According to the data set, many observations were made null if the total number of deaths during the collection period was between 1-9. The CDC notes this is due to being in accordance with NCHS confidentiality standards. Also, for the purpose of our project and report, we will only be using the data which summarizes the total values for the year and each month (only using observations 1-2755; the by total group). This is for the purpose of simplicity while also allowing us to work with a smaller (but still sufficient) data set size.

## Question we want to answer with this data?
We are using this data to answer what factors resulted in the most deaths to COVID, Influenza, and/or Pneumonia. In addition, we can use the fact that the data is separated by state to also analyze geographical influence/correlations as well.

## Outcome of interest?
A statistical representation/test on the significance of each variable on the number of deaths due to COVID, Influenza, and/or Pneumonia.

## Predictors?
The predictors we are focusing on are the following: geographical location, age, and sex.

## Relations/Patterns?
We are looking to see if there are any relationships in the data.

## Data Analysis Plan?
We plan to test the data using some sort of regression analysis or odds ratio to determine the correlation between predictors and outcome.

```{r}
# Loads the data into R. The original data is in the "CSV" format
raw_sa <- readr::read_csv(here::here("data", "raw_data", 
                                     "Provisional_COVID-19_Deaths_by_Sex_and_Age.csv"))
raw_deaths <- readr::read_csv(here::here("data", "raw_data", "deaths.csv"))
raw_gov <- readxl::read_xlsx(here::here("data", "raw_data", "government_22.xlsx"))
raw_econ <- readxl::read_xlsx(here::here("data", "raw_data",
                                         "economics_22.xlsx"),
                              sheet='per_capita_income')

# Provides an overview of the data and data types in the data frame.
dplyr::glimpse(raw_deaths)
dplyr::glimpse(raw_sa)
dplyr::glimpse(raw_gov)
dplyr::glimpse(raw_econ)
```

# Part 2
```{r}
# Includes only the data we intend to use for our analysis.
new_sa <- raw_sa %>%
# Identifies total values for Georgia subjects in the year 2020 only
              dplyr::filter(Year == 2020 & is.na(Month) & 
                            State == "Georgia" & 
# Filters out redundancies caused by overlapping age intervals
                            `Age Group` %in% c("Under 1 year","1-4 years","5-14 years","15-24 years",
                                               "25-34 years","35-44 years","45-54 years",
                                               "55-64 years","65-74 years","75-84 years",
                                               "85 years and over") &
# Filters out overlaps due to totals rows
                            Sex %in% c("Male","Female")) %>%
# Rids the dataset of "Notes" section and date variables
              dplyr::select(!c(Footnote,`Data As Of`,`Start Date`,`End Date`,Group,Year,Month,State))

# Checks that the data was manipulated properly
dplyr::glimpse(new_sa)
```

```{r}
# Removes unwanted "+" symbols from some rows
raw_deaths$age <- gsub('[^[:alnum:] ]','',raw_deaths$age)

# Converts the age variable to numeric and capitalizes the county variable
new_deaths <- raw_deaths %>%
                        dplyr::mutate(age = as.numeric(age),
                                      county = toupper(county))

# Checks that the data was manipulated properly
dplyr::glimpse(new_deaths)
```


```{r}
# Uses the voting proportions from 2020 to identify party majority in each county
new_gov <- raw_gov %>%
                   dplyr::mutate(party_majority = 
                                   ifelse(`2020 Votes Cast for President, Democratic Party, Percent`>
                                          `2020 Votes Cast for President, Republican Party, Percent`,
                                          "Democratic",
                                          "Republican")) %>%
                   dplyr::select(County,party_majority)

# Identifies the 2020 Per Capita Personal Income by county
new_econ <- raw_econ %>%
                     dplyr::select(County,`2020 Per Capita Personal Income`)
```


```{r}
# Joins the data from the `new_econ` data frame to the `new_deaths` data frame based on the county variable
join1 <- dplyr::left_join(new_deaths,new_econ,by=c("county"="County"))
# Checks that the join worked
dplyr::glimpse(join1)

# Joins the data from the `new_gov` data frame to the `join1` data frame based on the county variable
join2 <- dplyr::left_join(join1,new_gov,by=c("county"="County"))%>%
# Reorders the variables in the desired order
         dplyr::relocate(county,`2020 Per Capita Personal Income`,party_majority,.before=age)

# Checks that the join worked
glimpse(join2)
# Saves the cleaned data set and stores in in the processed_data folder
saveRDS(join2, here::here("data","processed_data","GACOVID_deaths"))
```


# References



